---
title: "SI2017_clustering_exercise_draft_v1"
author: "pfr"
date: "August 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## --------------------------------------------------
##    PFR data prep exercises for clustering
## NOTE:
##   Use the W_mncntr matrix from the first code chunks 
##   from the dataprep SVD dimension reduciotn exercise
##
##  **** If W_mncntr does not exist then go back and run those chunks ***
## 
## --------------------------------------------------



##select numeric columns
First, SVD and PCA only work on numeric columns,
so we have to only keep the numeric columns


```{r get numeric columns, echo=TRUE}


# Get numeric columns only
col_classes = sapply(W_df,class)   #get column classes as a list
num_inds    = c(which(col_classes=='numeric'), which(col_classes=='integer'))
                 #get column number indices 
W_dfnum       = W_df[,num_inds]
dim(W_dfnum)

```


##Run Kmeans for several K values


```{r kmeans, echo=TRUE}
#W_mncntrsc=scale(W_mncntr,center=TRUE,scale=TRUE)



# Run kmeans for 20 values of K
kg=matrix(0,20,2)
for (i in 1:20){
  ktest=kmeans(W_mncntr,i,20,5);
  kg[i,1]=i;
  kg[i,2]=ktest$tot.withinss;  #save total within Sum Square Dist to center
}
plot(kg[,1],kg[,2],main='kmeans within cluster SS, weatherdata numerics')

#try
str(ktest)
#for example:
#  cluster has a cluster number assignments for each datum
#  centers are center points for each cluster
#  tot sum of squares
#  within sum of squares

```

##Now, let's combine the Kmeans cluster information with the 
## SVD reduced dataset

## NOTE:  The data matrix W_dfred should exist, 
##     
##  *** If W_dfred is not available, otherwise go back to SVD
##      exercise and rerun that   ***


```{r combine Kmeans cluster with SVD info, echo=TRUE}

#get Kmeans for 4 clusters 
k4             = kmeans(W_mncntr,4,20,5)
#get color scheme
col2use        = c('red','blue','black','yellow')
#get cluster assignment in colors
colassignments = col2use[k4$cluster]

W_proj = as.matrix(W_mncntr) %*% Wsvd$v[,1:3]    #project data onto 3 components

plot(W_proj[,1],W_proj[,2],col=colassignments,main='data pts project to 1,2 SVD components, colored by kmeans')

# to plot center points, first project them into components
c3 = k4$centers%*% Wsvd$v[,1:3]
points(c3[,1],c3[,2],pch=8,cex=2)


```

## Now try coloring by class (using Y mean centered)


```{r plot with Y color}

#Y was created in SVD exercise, use it to select 2 colors

#get class assignment in colors
colassignments = col2use[Y]

plot(W_proj[,1],W_proj[,2],col=colassignments,main='data pts project to 1,2 SVD components, colored by class')




```



