---
title: "Data Exploration on Weather Data"
author: "Mai H. Nguyen"
date: "2017-07-20”
output: html_document
---

#### Preliminaries
##### Get data from rattle library
```{r getData, eval=TRUE, echo=TRUE}
library(rattle)
df <- weather

# If reading data from CSV file instead of loading from rattle package.
# df <- read.csv("weather-orig.csv",header=TRUE,row.names=FALSE)
```

##### Remove variable RISK_MM, which is the same as Rainfall for the next day
```{r  removeRiskMM, eval=TRUE, echo=TRUE}
df$RISK_MM <- NULL                               
```

#### Data Exploration
##### Check statistics and values in dataset
```{r stats, eval=TRUE, echo=TRUE}
str(df)             # Structure of object
dim(df)             # Dimensions of object
names(df)           # Column headers
head (df)           # First few lines of data
tail(df)            # Last few lines of data
summary(df)         # Summary statistics
```

##### Remove rows with missing values
```{r na, eval=TRUE, echo=TRUE}

# Find number of places in dataset where there are missing values
sum(is.na(df))

# Find number of samples (i.e., rows) with any missing value
sum(complete.cases(weather))

# Remove rows with any missing value
df <- na.omit(df)
dim(df)
summary(df)
```

##### Statistics on individual variables
```{r statsVar, eval=TRUE, echo=TRUE}

# Summary statistics
mean(df$MinTemp)
var(df$MinTemp)
sd(df$MinTemp)
summary(df$MinTemp)

# Other statistics
min(df$Date)                                # First date
max(df$Date)                                # Last date
sum(df$Rainfall)                            # Total rainfall amount for year
sum(df$Rainfall) / length(df$Rainfall)      # Average daily rainfall
```

##### Save the data as a CSV file (RISK_MM variable and NAs removed)
```{r saveCSV}
write.csv(df, "data/weather.csv", row.names=FALSE)
```

##### Explore numerical variables
```{r numVars, eval=TRUE, echo=TRUE}
# Numerical variables (full set)
# numeric.vars <- c("MinTemp","MaxTemp","Rainfall","Evaporation","Sunshine","WindGustSpeed","WindSpeed9am","WindSpeed3pm","Humidity9am","Humidity3pm","Pressure9am","Pressure3pm","Cloud9am","Cloud3pm","Temp9am","Temp3pm")

# Subset of numerical variables.
numeric.vars <- c("Rainfall","Evaporation","Sunshine","WindSpeed9am","Humidity9am","Pressure9am","Cloud9am","Temp9am")

# Create DF with just numeric variables
df.num <- df[numeric.vars]
dim(df.num)
head(df.num)
```

##### Compute and plot pairwise correlation matrix
```{r corr, evalL=TRUE, echo=TRUE}
cor(df.num, use="pairwise")
```

##### Explore categorical variables
```{r ctgVars, eval=TRUE, echo=TRUE}
# Non-numeric variables(only a subset is used)
ctg.vars <- c("Date","Location","WindGustDir","WindDir9am","WindDir3pm","RainToday","RainTomorrow")

# Levels (categories) for WindGustDir
levels(df$WindGustDir)

# Number of occurrences for each WindGustDir category
table(df$WindGustDir)

# Sorted number of occurrences for each level
sort(table(df$WindGustDir))

# Percentage for each WindGustDir category
round (table(df$WindGustDir) / sum(table(df$WindGustDir)) * 100, digits=2)

# Sorted percentages for WindGustDir categories
sort (round (table(df$WindGustDir) / sum(table(df$WindGustDir)) * 100, digits=2))
```

#### Plots for visualizing data
##### Histogram of Sunshine
```{r hist, eval=TRUE, echo=TRUE}
summary(df$Sunshine)
hist(df$Sunshine,col="lightblue",main="Histogram of Daily Sunshine",xlab="Sunshine (Hours)")
```

##### Density plot of Sunshine
```{r density, eval=TRUE, echo=TRUE}
plot(density(df$Sunshine), main="Distribution of Daily Sunshine (Hours)")
plot(density(df$Sunshine),ylim=c(0,0.2), main="Distribution of Daily Sunshine (Hours)")
```

##### Bar plot showing frequency of wind gust directions
```{r barplot, eval=TRUE, echo=TRUE}
table(df$WindGustDir)
plot(df$WindGustDir,col="lightblue",main="Distribution of Wind Gust Direction",xlab="WindGustDir",ylab="Count")
```

##### Scatter plot of Evaporation vs. Temperature
```{r scatterplot, eval=TRUE, echo=TRUE}
cor(df$Evaporation,df$Temp9am)
plot(df$Evaporation,df$Temp9am,col="blue",main="Evaporation vs. Temperature",xlab="Evaporation (mm)",ylab="Temp9am (C)")
```

##### Scatter plot of Evaporation vs. Temperature with regression line
```{r scatterplot2, eval=TRUE, echo=TRUE}
plot(df$Evaporation,df$Temp9am,col="blue",main="Evaporation vs. Temperature",xlab="Evaporation (mm)",ylab="Temp9am (C)")
abline(lm(df$Temp9am ~ df$Evaporation), col="red")
```

##### Line plot of daily high temperature
```{r lineplot, eval=TRUE, echo=TRUE}
summary(df$MaxTemp)

# Using base R plot command
plot(df$Date,df$MaxTemp,type='l',col="blue",main="Daily High Temperature",xlab="Date",ylab="MaxTemp (C)")

# Using ggplot2 package
library(ggplot2)     # R package for plotting that uses grammar of graphics
ggplot(df, aes(x=Date,y=MaxTemp)) + geom_line(color="blue") + geom_smooth(color="red",size=1) + ggtitle("Daily High Temperature") + ylab("MaxTemp (C)")
```

##### Box plot of atmospheric pressure with respect to RainToday
```{r boxplot, eval=TRUE, echo=TRUE}
ggplot(df, aes(x=RainToday, y=Pressure9am, fill=RainToday)) + geom_boxplot() +
  ggtitle("Atmospheric Pressure wrt Rain") 
```

##### Box plot of MaxTemp by month
```{r boxplot2, eval=TRUE, echo=TRUE}
library(lubridate)   # R package for handling dates
ggplot(df, aes(x=as.factor(month(Date)), y=MaxTemp)) + geom_boxplot(fill="lightblue") + 
  xlab("Month") + ylab("MaxTemp (Celsius)") + ggtitle("High Temperature by Month")
```

##### Violin plot of MaxTemp by month
```{r violinplot, eval=TRUE, echo=TRUE}
ggplot(df, aes(x=as.factor(month(Date)), y=MaxTemp)) + geom_violin(fill="lightblue") + 
  xlab("Month") + ylab("MaxTemp (Celsius)") + ggtitle("High Temperature by Month")
```

##### Plot correlation matrix
```{r corrplot, evalL=TRUE, echo=TRUE}
library(GGally)        # R package that extends ggplot2  
cor(df.num, use = "pairwise")
ggcorr(df.num, use="pairwise", label=TRUE, label_round=2)  # ggcorr uses cor to calculate correlations
```

##### Plot pairwise correlation
```{r pairsplot, eval=TRUE, echo=TRUE}
ggpairs(df[c("Sunshine","Pressure9am","Humidity9am","Temp9am")])  # ggpairs is from GGally package
```